Imprecise Datapath Design: An Overclocking Approach

### 2014-01-17 TRETS2013/ReviewResponse/Response.tex

\documentclass[a4paper, 11pt]{article}

\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{xcolor}

\topmargin -10mm
\oddsidemargin 0mm
\evensidemargin 0mm
\textwidth 158mm
\textheight 226mm
\parskip 0mm

\def\Response{\noindent \textbf{Response:~}}
\newcommand{\Question}[1]{\textcolor[rgb]{0.51,0.00,0.00}{#1}}
\newcommand{\PaperText}[1]{\emph{#1}}

\title{Revision of Paper TRETS-2013-0065 Imprecise Datapath Design: An Overclocking Approach}
\author{Kan Shi, David Boland and George A. Constantinides}
\date{}

\pagenumbering{arabic}

\begin{document}
\maketitle

\section*{Reviewer 3}
\begin{enumerate}
  \item \Question{The first sentence of the abstract is completely devoid of context, and may be the worst sentence in the entire article, given its importance. A much more appropriate sentence is the start of Section 1, Paragraph 4.}

      \Response We have modified the first two sentences of the abstract, which now read as follows:\\

      \PaperText{In this paper, we describe an alternative circuit design methodology when considering trade-offs between accuracy, performance and silicon area. We compare two different approaches that could trade accuracy for performance.}\\

  \item \Question{What are the chances that an entire design with all its complexities could be accelerated in this fashion? (i.e. Could the overall clock frequency be increased?)}

      \Response to be added.\\

  \item \Question{In Figure 1, you use $\mu$, whereas in the text and in Equation 4, you use $\mu_c$.  The converse is true for Section 7.4 and equations (35), (36), and (37).}

      \Response Thanks. It should be $\mu$ in Fig.1 and Equations (35), (36) and (37). It has been fixed.\\

  \item \Question{In Section 3.2, it would be helpful if you could illustrate or define your [-1,1) fixed-point format. Do you number your bits from k-1 to 0 or from 0 to k-1?}

      \Response We meant that all numbers are normalized to fractional numbers within the range [-1,1). The truncation process is illustrated in the figure below, which is not included in this manuscript due to the space limit. As seen in this figure, the original number embodies k fractional bits, while the truncated number embodies n fractional bits. Given that all bits are mutually independent and uniformly distributed, Equation (2) gives the mean value of the truncated bits. We have modified the second paragraph of Section 3.2:\\

      \PaperText{to be added.}\\

  \item \Question{In Section 3.3, for $A_i\neq B_i$ (Carry Propagation), you seem to be saying that $S_i = C_{i-1} = 0$.  If $A_i \neq B_i$ then $A_i + B_i = 1$, but that tells us nothing unless we know $C_{i-1}$, and it is not necessarily 0 as you claim.}

      \Response Thanks for pointing this out. In this case it should be $C_{i-1} =1$, because carry is propagated from bit $i-1$ to bit $i$. Hence $S_i=A_i+B_i+C_{i-1}=0$.

      This statement is modified as follows:\\

      \PaperText{Carry propagation: (see paper)}

\end{enumerate}

\end{document}

### 2014-01-20 TRETS2013/TRETS_overclocking.tex
Reviewer 2: Main body

% Package to generate and customize Algorithm as per ACM style
\usepackage[ruled]{algorithm2e}
%\usepackage{amsmath,epsfig}
%\usepackage[table]{xcolor}
%\usepackage{amsmath,amsfonts,epsfig}
%\usepackage{fancyhdr}
%\usepackage{mathdots,colortbl,dcolumn,rotating,subfigure,appendix}
%\usepackage[scriptsize,it]{caption}
\usepackage{graphicx}
%\usepackage{epstopdf}
\usepackage{multirow}
\usepackage{amssymb}
\usepackage[cmex10]{amsmath}
\usepackage[tight,footnotesize]{subfigure}
\usepackage{stfloats}
\usepackage{threeparttable}

@@

\begin{abstract}
%As process scaling introduces significant performance variations, much more conservative timing margins will have to be applied to avoid timing violations. Potentially, r
Releasing the tight accuracy requirement would potentially offer greater freedom to create a design with better performance or energy efficiency. In this paper, we compare two different approaches that could trade accuracy for performance. One is the traditional approach where the precision used in the datapath is limited to meet a target latency. The other is a proposed new approach which simply allows the datapath to operate without timing closure. We demonstrate analytically and experimentally that on average our approach obtains either smaller errors for equivalent faster operating frequencies in comparison to the traditional approach. This is because the worst case caused by timing violations only happens rarely, while precision loss results in errors to most data. We also show that for basic arithmetic operations such as addition, applying our approach to the simple building block of ripple carry adders can achieve better accuracy or performance than using faster adder designs to achieve similar latency.

 %operators (i.e. ripple carry adders) can achieve better accuracy or performance than using the traditional method to both original operators and intrinsically faster operators (i.e. carry select adders) which consumes the same silicon area.
\end{abstract}

\category{C.2.2}{Category put here}{Detail category}

\terms{Design, FPGA, Performance}

\keywords{Overclocking, Imprecise Design, Approximate Computing, Truncation}

@@

Performance is a critical issue when designing embedded computing systems. Specifically, a large volume of current studies has demonstrated that, significant performance gains can be achieved by FPGA-based accelerators over software designs across a wide range of applications~\cite{UnderwoodFPGAvsCPU,David_MINRES}. However, one of the major factors that limits the performance of these accelerators is that they typically run at much lower clock frequencies than general purpose processors (GPPs) or graphics processing units (GPUs).

@@

In this paper, we describe an alternative circuit design methodology when considering trade-offs between accuracy, performance and silicon area. Since it is unlikely to create a completly error-free design as any data representation would introduce quantization errors, we take a radical shift forward by suggesting that for certain applications it is beneficial to move away from the traditional model of creating a conservative design that is guaranteed to avoid timing violations. Instead, it may be preferable to create a design in which timing violations may occur, under the knowledge that they only occur rarely because specific input patterns are required to generate the worst case errors. This paper elaborates on the prior work~\cite{SKfccm13}, by incorporating silicon area as another evaluation metric. This enables us to find the optimum design choice of the basic arithmetic operators under various design constraints. A summary of the main contributions of this work are as follows:

\begin{itemize}
	\item Detailed modeling methods of probabilistic errors generated in basic arithmetic primitives from two design scenarios to trade accuracy for performance: overclocking the datapath or truncating precisions that used throughout the datpath,
	\item Exploring the accuracy-performance-area trade-offs for key arithmetic operators to illustrate the optimum architecture for a certain arithmetic function under a given requirement of silicon area,
	\item Demonstration of the proposed findings analytically and experimentially from FPGA implementations that allowing rare timing violations to happen leads to a reduction of average errors or an improvement of performance over trancating the datapath precisions to meet timing.
\end{itemize}

The rest of this paper is organized as follows. It first reviews in detail the current literature regarding the approaches used in word-length optimization and approximate datapath design in Section~\ref{Section_Background}. The analytical error models for two different adder structures: the ripple carry adder (RCA) and the carry-select adder (CSA), are discussed in Section~\ref{section_RCA} and Section~\ref{Section_CSA}, respectively. The decision of the optimum adder structure under the accuracy-performance-area trade-offs, is then put forward in Section~\ref{Section_OptimumAdder}. It is followed by a description of the probabilistic error models for another arithmetic operator, the constant coefficient multiplier (CCM), in Section~\ref{section_CCM}. Section~\ref{Section_ExpSetup} discusses a practical experimental setup to verify our models and test the proposed design methodology. We then demonstrate the benefits of our proposed approach in Section~\ref{Section_Experiments}. Finally, conclusion of this work and possible future work are discussed in Section~\ref{Section_Conclusion}. 
%discusses the conclusions of this work and the possible research directions in the future.

@@

According to the existing literature, one way to operate circuits beyond the deterministic region is to relax the design constraints and the guard-bands that are conventionally used to avoid the worst cases. A series of work known as ``Better-Than-Worst-Case (BTWC) Design'' introduced a universal framework to push the circuit performance to its limits~\cite{BetterThanWS2005}. In general, a BTWC design is composed of cores and checkers. The cores are operated with high performance by eliminating the guard-bands, meanwhile the possible timing errors are diagnosed by the checkers. Optionally the system can be recovered at the observation of errors. As an exemplary design, the Razor project~\cite{Razor2003,Razor2004} was proposed to shave the conservative timing margins by overscaling the supply voltage and clock frequency, while monitoring the output error rates by utilizing a self-checking circuit. This work demonstrated that the benefits brought by removing the safe margin outweigh the cost of monitoring and recovering from errors. For example, $22\%$ or over $30\%$ power consumption can be saved with $\sim0.01\%$ or $\sim1\%$ error rates at the output, respectively. Related work also includes a similar frequency overscaling technique by operating circuits slightly slower than the critical path delay with dedicated checker circuits to ensure that timing errors will not occur~\cite{TEAtime2004}, or developing timing analysis tools that decide the optimum operating frequencies in the non-deterministic region due to process variation~\cite{ProbabilisticSTA}. However, the major problem of the BTWC work is the extra cost introduced by implementing the checker circuits.
%of silicon area due to the implementation of the checker circuits.

@@

While the aforementioned literature take advantage of the fact that only specific input patterns could cause timing errors, research on imprecise architectures take this one step further by designing simplified circuits for peformance and/or energy efficiency. For example, Lu et~al. proposed a ``shrinking'' datapath that can be utilized to mimic and speculate the original logic functions~\cite{IntelSpeeding}. Kulkarni et.al. described an underdesigned $2\times2$ multiplier unit, of which the worst case was replaced by a normal case based on the straight-forward Karnaugh-Map analysis~\cite{Undersigned2x2multiplier}. In both cases, reduction of area and power consumption can be achieved with the cost of accuracy. In addition, Gupta et~al. developed approximate adders at the transistor level and compared the energy efficiency of their proposed architectures over truncation of input word-length of conventional structures~\cite{Gupta2013TransCADICS}. However, we should note that using imprecise architectures means correct results would never be obtained for certain input patterns. Another limitation is that the link between the probability of output correctness and energy savings or performance improvements are rarely analyzed in current research. In addition, these techniques cannot be directly applied onto existing hardware platforms such as FPGAs.

@@

%and  that are commonly used in DSP applications and numerical algorithms.

@@

% $\mu_{RCA}=n\mu$, as shown in~. 
For an $n$-bit RCA, it follows that if the sampling period $T_S$ is greater than $\mu_{RCA}$, correct results will be sampled. If, however, faster-than-rated sampling frequencies are applied such that $T_S<\mu_{RCA}$, intermediate results will be sampled, potentially generating errors.

\begin{figure}[t]
  \centering
  \includegraphics[width=3in]{./Figures/FastCarryLogic3.eps}
  %  \vspace{-2ex}
  \caption{An $n$-bit ripple carry adder in the Xilinx Virtex-6 FPGA.}
  \label{FPGA adder}
  %\vspace{-3.5ex}
\end{figure}

@@

However, since the length of an actual carry chain during execution is dependent upon input patterns, in general, the worst case may occur rarely. To determine when this timing constraint is not met and the size of the error in this case, we expand standard results~\cite{DigitalICDesign} to the following statements, which examine carry generation, propagation and annihilation, as well as the corresponding summation results of a single bit $i$ within an $n$-bit RCA ($0\leq i<n$), according to the relationship between its input patterns $A_i$ and $B_i$:
%
 \begin{itemize}
   \item Carry generation: $A_i=B_i=1$, a new carry chain is generated at bit $i$, and $S_i=C_{i-1}$;
   \item Carry propagation: $A_i\neq B_i$, the carry propagates for this carry chain at bit $i$, and $S_i=\bar{C_{i-1}}=0$;
   \item Carry annihilation: $A_i=B_i$ (either $0$ or $1$), the current carry chain annihilates at bit $i$, and $S_i=C_{i-1}=1$.
 \end{itemize}

@@

For $C_{tm}$, correct results will be generated from bit $S_t$ to bit $S_{t+b-1}$. Hence the absolute value of error seen at the output, normalized to the MSB ($2^n$), is given by (\ref{Eq:Output_Error}), where $\hat{S_i}$ and $S_i$ denote the actual value and error-free value of outputs at bit $i$, respectively.
%
\begin{eqnarray}\label{Eq:Output_Error}
%\small
%\footnotesize
    e_{tm}=\frac{\left|\sum_{i=t+b}^{n}(S_i-\hat{S_i})\cdot 2^i\right|}{2^n}
%    \vspace{-1ex}
\end{eqnarray}

@@

In addition, note that the probabilistic models for the magnitude and probability of overclocking errors can be easily generalized for any input distributions with mutually independent bits, since (\ref{Eq:Output_Error}) and (\ref{Eq:etm}) are not related to the input distribution, while (\ref{Eq:Ptm_RCA}) can be employed as long as the value of $P_{(A_i=B_i=1)}$ and $P_{(A_i\neq B_i)}$ is known. If there is dependency between input bits, (\ref{Eq:Ptm_RCA_General}) should be used and the probability value of $P_{(A_i=B_i=1)}$ and $P_{(A_i\neq B_i)}$ would change. For instance, if $A_i$ and $B_i$ are not independent, we have $P_{(A_i=1)}=E_{(A_i)}=1/2$ and $P_{(B_i=1)}=E_{(B_i)}=1/2$ seperately for uniform distribution, where $E_{(A_i)}$ denotes the expectation of $A_i$. The probability of $A_i=B_i=1$ is then given by (\ref{Eq:P_dependency})~\cite{TransitionProb} where $\sigma_{(A_i)}$ denotes the standard variation of $A_i$ and $\rho_i$ denotes the temporal correlation between the two inputs at the $i^{th}$ bit. $\rho_i=0$ corresponds to the case where bits are uncorrelated. 
%
\begin{eqnarray}\label{Eq:P_dependency}
    P_{(A_i=B_i=1)}=E_{(A_i)}E_{(B_i)}+\rho_i\sigma_{(A_i)}\sigma_{(B_i)}
\end{eqnarray}

@@

\section{Carry Select Adder}\label{Section_CSA}
\subsection{Introduction}
As seen in~(\ref{Eq:Comparison RCA}), a factor of 2 reduction in error expection can be achieved by using the proposed approach in RCA. However, implementing our approach costs extra silicon area in comparison to the conventional approach, because we keep the original word-length $k$ in the new scenario instead of truncating to only $b-1$ bits in the traditional scenario. Besides our approach, there are existing approaches that could be employed to trade silicon area for low latency. For instance, alternative adder architectures, such as carry select adder (CSA), have been proposed to boost performance. In a CSA, the carry chain is divided into multiple overlapped stages, while each stage contains two RCAs and two multiplexers, as shown in Fig.~\ref{Fig_CSA_Structure} and Fig.~\ref{Fig_CSA_SingleStage}, respectively. For a given input, two additions are performed simultaneously within a single stage where the carry input is zero and one seperately. One of these two results is then selected according to the actual carry input. Although this structure brings performance benefits, it costs extra hardware resources compared to a standard RCA because the carry chain is duplicated. Furthermore, in FPGA technology, multiplexers are expensive to implement. Due to this reason, we explore the trade-offs between silicon area, accuracy and performance of RCA and CSA in this section.
\vspace{-7ex}
\begin{figure}[htbp]
        \centering
        \subfigure[A CSA with $s$ stages.]{
            \includegraphics[width=0.6\textwidth]{./Figures/CSA_structure.pdf}
            \label{Fig_CSA_Structure}
        }%\vspace{-2ex}
        \subfigure[The $i^{th}$ stage in the CSA.]{
        \centering
            \includegraphics[width=2.2in]{./Figures/CSA_SingleStage.pdf}
            \label{Fig_CSA_SingleStage}
        }
    \vspace{-2ex}
    \caption{The structure diagram of CSA.}
\end{figure}
\vspace{-2ex}

@@

Fig.~\ref{Fig_CSA2stage_Timing} also highlights that for a relatively relexed frequency cosntraint, the CSA achieves greater word-length than the RCA. This is because the stage parallelism in the CSA enables a greater overall word-length, even though the multiplexer delay limits the word-length of each stage in the CSA when compared to the RCA. When more stringent latency requirement is applied, however, the word-length of each stage is largely limited such that the multiplexer delay becomes comparable to the delay of the carry chain, and this inhibits the benefits of parallelism.

%in comparison to the RCA, the CSA achieves greater word-length for relatively relexed frequency constraints. The RCA only outperforms when very high frequency requirement is applied. This is because at low frequencies, although the multiplexer delay limits the word-length of each stage in CSA when compared to RCA, the stage parallelism in CSA enables a greater word-length. However when more stringent latency requirement is applied, the word-length of each stage in CSA is limited. Hence the multiplexer delay becomes comparable to the delay of the carry chain, and this inhibits the benefits of parallelism.

@@

As an example slice through the design space, in Fig.~\ref{Fig_CSA_LUT25} we record the mean value of error at outputs with respect to different operating frequencies for both design scenarios when the number of the available LUTs is set to 25. In this graph, we have labeled the optimum adder structure. Note that in this and all the following experiments within this section, in order to apply our models the input data is randomly generated following the uniform distribution.
%
\begin{figure}[t]
  \centering
  \vspace{-3ex}
  \includegraphics[width=3.3in]{./Figures/Error_LUT25.eps}
  \vspace{-2ex}
  \caption{A comparison between two design scenarios when the number of available LUT is 25. The RCA and CSA are investigated in the conventional scenario, while the RCA is explored in the proposed new scenario. The results are obtained from post place-and-route simulations on Xilinx Virtex-6 FPGAs.}
  \vspace{-.5ex}
  \label{Fig_CSA_LUT25}
\end{figure}

We first notice from Fig.~\ref{Fig_CSA_LUT25} that for all frequency values, the overclocked RCA achieves no larger error expectation than the RCA with truncated operand word-lengths, as predicted by the models for the RCA in Section~\ref{section_RCA_Comparison}. It can also be observed that the CSA cannot be implemented with the original word-length due to the limited area budget and this leads to large truncation errors. Although the CSA with 4 stages is best for some frequencies, the overclocked RCA is still the optimum design when high operating frequencies are applied.

We then perform similar experiments with a variety of area constraints. The optimum design methods with respect to different operating frequencies and area consumptions are demonstrated in Fig.~\ref{Fig_CSA_Tradeoff}. From this figure, several observations can be made. Firstly, if the frequency requirement is not large whilst the available area is large enough to implement a CSA in full precision, it will be the optimum design. This is expected from our earlier analysis in Fig.~\ref{Fig_CSA2stage_Timing}. Secondly, the 2-stage CSA is better than the 3-stage CSA and the 4-stage CSA when frequency is low, as it consumes less area although all achieve the same error expectation. Thirdly, it is possible that only part of the CSA can be implemented under a tighter area budget, whereas the RCA still keeps full precision. In this case, area becomes the dominant factor and precision is lost for the CSA, meaning the RCA with overclocking is the optimum design method across almost the whole frequency domain in this situation. Finally, for very stringent area constraints, the word-length of RCA is also limited. This results in truncation errors initially for all design scenarios. However, the RCA with overclocking can still be employed as the optimum design, because it loses less precision than the CSA under the same area constraint.
%
\begin{figure}[t]
  \centering
  \vspace{-2.5ex}
  \includegraphics[width=4.5in]{./Figures/Tradeoff_Freq.eps}
  \vspace{-2.5ex}
  \caption{A demonstration of the optimum design methodology which achieves the minimum error at outputs with respect to a variety of frequency and area constraints.}
  \label{Fig_CSA_Tradeoff}
\end{figure}

\subsection{Determination of the Optimum Adder Structure for Given Accuracy Requirements}
If the design goal is to operate the circuit as fast as possible with the minimum area whilst a certain error budget can be tolerated, the optimum design methodology can be decided as illustrated in Fig.~\ref{Fig_CSA_Tradeoff_Error}. In this situation, the error specifications are evaluated in terms of the mean relative error ($MRE$), as given by~(\ref{Eq:MRE}), where $E_{error}$ and $E_{out}$ refer to the mean value of error and the mean value of outputs, respectively.
%
\begin{eqnarray}\label{Eq:MRE}
  MRE=\left|\frac{E_{error}}{E_{out}}\right|\times 100\%
\end{eqnarray}

@@

%In general, we can observe a relatively small area overhead of our approach, especially

@@

%explore the probabilistic behaviour of alternative

>>>>

% Package to generate and customize Algorithm as per ACM style
\usepackage[ruled]{algorithm2e}
%\usepackage{amsmath,epsfig}
%\usepackage[table]{xcolor}
%\usepackage{amsmath,amsfonts,epsfig}
%\usepackage{fancyhdr}
%\usepackage{mathdots,colortbl,dcolumn,rotating,subfigure,appendix}
%\usepackage[scriptsize,it]{caption}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{multirow}
\usepackage{amssymb}
\usepackage[cmex10]{amsmath}
\usepackage[tight,footnotesize]{subfigure}
\usepackage{stfloats}
\usepackage{threeparttable}

@@

\begin{abstract}
%Releasing the tight accuracy requirement would potentially offer greater freedom to create a design with better performance or energy efficiency.
In this paper, we describe an alternative circuit design methodology when considering trade-offs between accuracy, performance and silicon area. We compare two different approaches that could trade accuracy for performance. One is the traditional approach where the precision used in the datapath is limited to meet a target latency. The other is a proposed new approach which simply allows the datapath to operate without timing closure. We demonstrate analytically and experimentally that on average our approach obtains either smaller errors for equivalent faster operating frequencies in comparison to the traditional approach. This is because the worst case caused by timing violations only happens rarely, while precision loss results in errors to most data. We also show that for basic arithmetic operations such as addition, applying our approach to the simple building block of ripple carry adders can achieve better accuracy or performance than using faster adder designs to achieve similar latency.
\end{abstract}

\category{B.3.3}{Performance Analysis and Design Aids}{Detail category}

\terms{Design Methodologies, Overclocking, FPGA, Performance Modeling and Analysis}

\keywords{Approximate Computing, Truncation, Imprecise Design}

@@

Performance is a critical issue when designing embedded computing systems. Specifically, a large volume of current studies has demonstrated that significant performance gains can be achieved by FPGA-based accelerators over software designs across a wide range of applications~\cite{UnderwoodFPGAvsCPU,David_MINRES}. However, one of the major factors that limits the performance of these accelerators is that they typically run at much lower clock frequencies than general purpose processors (GPPs) or graphics processing units (GPUs).

@@

In this paper, we describe an alternative circuit design methodology when considering trade-offs between accuracy, performance and silicon area. Since it is unlikely to create a completely error-free design as any data representation would introduce quantization errors, we take a radical shift forward by suggesting that for certain applications it is beneficial to move away from the traditional model of creating a conservative design that is guaranteed to avoid timing violations. Instead, it may be preferable to create a design in which timing violations may occur, under the knowledge that they only occur rarely because specific input patterns are required to generate the worst case errors. This paper elaborates on the prior work~\cite{SKfccm13}, by incorporating silicon area as another evaluation metric. This enables us to find the optimum design choice of the basic arithmetic operators under various design constraints. A summary of the main contributions of this work are as follows:

\begin{itemize}
	\item Detailed modeling methods of probabilistic errors generated in basic arithmetic primitives from two design scenarios to trade accuracy for performance: overclocking the datapath or truncating precisions used throughout the datapath,
	\item Exploring the accuracy-performance-area trade-offs for key arithmetic operators to illustrate the optimum architecture for a certain arithmetic function under a given requirement of silicon area,
	\item Demonstration of the proposed findings analytically and experimentally from FPGA implementations that allowing rare timing violations to happen leads to a reduction of average errors or an improvement of performance over truncating the datapath precisions to meet timing.
\end{itemize}

The rest of this paper is organized as follows: It first reviews in detail the current literature regarding the approaches used in word-length optimization and approximate datapath design in Section~\ref{Section_Background}. The analytical error models for two different adder structures: the ripple carry adder (RCA) and the carry-select adder (CSA), are discussed in Section~\ref{section_RCA} and Section~\ref{Section_CSA}, respectively. The decision of the optimum adder structure under the accuracy-performance-area trade-offs, is then put forward in Section~\ref{Section_OptimumAdder}. It is followed by a description of the probabilistic error models for another arithmetic operator, the constant coefficient multiplier (CCM), in Section~\ref{section_CCM}. Section~\ref{Section_ExpSetup} discusses a practical experimental setup to verify our models and test the proposed design methodology. We then demonstrate the benefits of our proposed approach in Section~\ref{Section_Experiments}. Finally, conclusion of this work and possible future work are discussed in Section~\ref{Section_Conclusion}.
%discusses the conclusions of this work and the possible research directions in the future.

@@

According to the existing literature, one way to operate circuits beyond the deterministic region is to relax the design constraints and the guard-bands that are conventionally used to avoid the worst cases. A series of work known as ``Better-Than-Worst-Case (BTWC) Design'' introduced a universal framework to push the circuit performance to its limits~\cite{BetterThanWS2005}. In general, a BTWC design is composed of cores and checkers. The cores are operated with high performance by eliminating the guard-bands, meanwhile the possible timing errors are diagnosed by the checkers. Optionally the system can be recovered at the observation of errors. As an exemplary design, the Razor project~\cite{Razor2003,Razor2004} was proposed to shave the conservative timing margins by overcalling the supply voltage and clock frequency, while monitoring the output error rates by utilizing a self-checking circuit. This work demonstrated that the benefits brought by removing the safe margin outweigh the cost of monitoring and recovering from errors. For example, over $30\%$ power consumption can be saved with $\sim1\%$ error rates at the output, respectively. Related work also includes a similar frequency overcalling technique by operating circuits slightly slower than the critical path delay with dedicated checker circuits to ensure that timing errors will not occur~\cite{TEAtime2004}, or developing timing analysis tools that decide the optimum operating frequencies in the non-deterministic region due to process variation~\cite{ProbabilisticSTA}. However, the major problem of the BTWC work is the extra cost introduced by implementing the checker circuits.
%of silicon area due to the implementation of the checker circuits.

@@

While the aforementioned literature take advantage of the fact that only specific input patterns could cause timing errors, research on imprecise architectures take this one step further by designing simplified circuits for performance and/or energy efficiency. For example, Lu et~al. proposed a ``shrinking'' datapath that can be utilized to mimic and speculate the original logic functions~\cite{IntelSpeeding}. Kulkarni et~al. described an under-designed $2\times2$ multiplier unit, of which the worst case was replaced by a normal case based on the straight-forward Karnaugh-Map analysis~\cite{Undersigned2x2multiplier}. In both cases, reduction of area and power consumption can be achieved with the cost of accuracy. In addition, Gupta et~al. developed approximate adders at the transistor level and compared the energy efficiency of their proposed architectures over truncation of input word-length of conventional structures~\cite{Gupta2013TransCADICS}. However, we should note that using imprecise architectures means correct results would never be obtained for certain input patterns. Another limitation is that the link between the probability of output correctness and energy savings or performance improvements are rarely analyzed in current research. In addition, these techniques cannot be directly applied onto existing hardware platforms such as FPGAs.

@@

@@

For an $n$-bit RCA, it follows that if the sampling period $T_S$ is greater than $\mu_{RCA}$, correct results will be sampled. If, however, faster-than-rated sampling frequencies are applied such that $T_S<\mu_{RCA}$, intermediate results will be sampled, potentially generating errors.

\begin{figure}[t]
  \centering
  \includegraphics[width=3.2in]{./Figures/FastCarryLogic.pdf}
  %  \vspace{-2ex}
  \caption{An $n$-bit ripple carry adder in the Xilinx Virtex-6 FPGA.}
  \label{FPGA adder}
  %\vspace{-3.5ex}
\end{figure}

@@

However, since the length of an actual carry chain during execution is dependent upon input patterns, in general, the worst case may occur rarely. To determine when this timing constraint is not met and the size of the error in this case, we expand standard results~\cite{DigitalICDesign} to the following statements, which examine carry generation, propagation and annihilation, as well as the corresponding summation results of a single bit $i$ within an $n$-bit RCA ($0\leq i<n$), according to the relationship between its input patterns $A_i$ and $B_i$:
%
 \begin{itemize}
   \item Carry generation: $A_i=B_i=1$, a new carry chain is generated at bit $i$, and $S_i=C_{i-1}$;
   \item Carry propagation: $A_i\neq B_i$, the carry propagates for this carry chain at bit $i$, and $\bar{C_{i-1}}=1$, $S_i=0$;
   \item Carry annihilation: $A_i=B_i$ (either $0$ or $1$), the current carry chain annihilates at bit $i$, and $S_i=C_{i-1}=1$.
 \end{itemize}

@@

For $C_{tm}$, correct results will be generated from bit $S_t$ to bit $S_{t+b-1}$. Hence the absolute value of error seen at the output, normalized to the MSB, is given by (\ref{Eq:Output_Error}), where $\hat{S_i}$ and $S_i$ denote the actual value and error-free value of outputs at bit $i$, respectively.
%
\begin{eqnarray}\label{Eq:Output_Error}
%\small
%\footnotesize
    e_{tm}=\frac{\left|\sum_{i=t+b}^{n}(S_i-\hat{S_i})\cdot 2^i\right|}{2^n}
%    \vspace{-1ex}
\end{eqnarray}

@@

In addition, note that the probabilistic models for the magnitude and probability of overclocking errors can be easily generalized for any input distributions with mutually independent bits, since (\ref{Eq:Output_Error}) and (\ref{Eq:etm}) are not related to the input distribution, while (\ref{Eq:Ptm_RCA}) can be employed as long as the value of $P_{(A_i=B_i=1)}$ and $P_{(A_i\neq B_i)}$ is known. If there is dependency between input bits, (\ref{Eq:Ptm_RCA_General}) should be used and the probability value of $P_{(A_i=B_i=1)}$ and $P_{(A_i\neq B_i)}$ would change. For instance, if $A_i$ and $B_i$ are not independent, we have $P_{(A_i=1)}=E_{(A_i)}=1/2$ and $P_{(B_i=1)}=E_{(B_i)}=1/2$ separately for uniform distribution, where $E_{(A_i)}$ denotes the expectation of $A_i$. The probability of $A_i=B_i=1$ is then given by (\ref{Eq:P_dependency})~\cite{TransitionProb} where $\sigma_{(A_i)}$ denotes the standard variation of $A_i$ and $\rho_i$ denotes the temporal correlation between the two inputs at the $i^{th}$ bit. $\rho_i=0$ corresponds to the case where bits are uncorrelated.
%
\begin{eqnarray}\label{Eq:P_dependency}
    P_{(A_i=B_i=1)}=E_{(A_i)}E_{(B_i)}+\rho_i\sigma_{(A_i)}\sigma_{(B_i)}
\end{eqnarray}

@@

\section{Carry Select Adder}\label{Section_CSA}
\subsection{Introduction}
As seen in~(\ref{Eq:Comparison RCA}), a factor of 2 reduction in error expectation can be achieved by using the proposed approach in RCA. However, implementing our approach costs extra silicon area in comparison to the conventional approach, because we keep the original word-length $k$ in the new scenario instead of truncating to only $b-1$ bits in the traditional scenario. Besides our approach, there are existing approaches that could be employed to trade silicon area for low latency. For instance, alternative adder architectures, such as carry select adder (CSA), have been proposed to boost performance. In a CSA, the carry chain is divided into multiple overlapped stages, while each stage contains two RCAs and two multiplexers, as shown in Fig.~\ref{Fig_CSA_Structure} and Fig.~\ref{Fig_CSA_SingleStage}, respectively. For a given input, two additions are performed simultaneously within a single stage where the carry input is zero and one separately. One of these two results is then selected according to the actual carry input. Although this structure brings performance benefits, it costs extra hardware resources compared to a standard RCA because the carry chain is duplicated. Furthermore, in FPGA technology, multiplexers are expensive to implement. Due to this reason, we explore the trade-offs between silicon area, accuracy and performance of RCA and CSA in this section.
\vspace{-7ex}
\begin{figure}[htbp]
        \centering
        \subfigure[A CSA with $s$ stages.]{
            \includegraphics[width=0.6\textwidth]{./Figures/CSA_structure.pdf}
            \label{Fig_CSA_Structure}
        }%\vspace{-2ex}
        \subfigure[The $i^{th}$ stage in the CSA.]{
        \centering
            \includegraphics[width=2.2in]{./Figures/CSA_SingleStage.pdf}
            \label{Fig_CSA_SingleStage}
        }
    \vspace{-2ex}
    \caption{The structure diagram of CSA.}
\end{figure}
\vspace{-2ex}

@@

Fig.~\ref{Fig_CSA2stage_Timing} also highlights that for a relatively relaxed frequency constraint, the CSA achieves greater word-length than the RCA. This is because the stage parallelism in the CSA enables a greater overall word-length, even though the multiplexer delay limits the word-length of each stage in the CSA when compared to the RCA. When more stringent latency requirement is applied, however, the word-length of each stage is largely limited such that the multiplexer delay becomes comparable to the delay of the carry chain, and this inhibits the benefits of parallelism.

@@

As an example slice through the design space, in Fig.~\ref{Fig_CSA_LUT25} we record the mean value of error at outputs with respect to different operating frequencies for both design scenarios when the number of the available LUTs is set to 25. In this graph, we have labeled the optimum adder structure. Note that in this and all the following experiments within this section, in order to apply our models the input data is randomly generated following the uniform distribution.
%
\begin{figure}[t]
  \centering
  \vspace{-3ex}
  \includegraphics[width=3.3in]{./Figures/Error_LUT25.eps}
  \vspace{-2ex}
  \caption{A comparison between two design scenarios when the number of available LUT is 25. The RCA and CSA are investigated in the conventional scenario, while the RCA is explored in the proposed new scenario. The results are obtained from post place-and-route simulations on Xilinx Virtex-6 FPGAs.}
  \vspace{-2ex}
  \label{Fig_CSA_LUT25}
\end{figure}

We first notice from Fig.~\ref{Fig_CSA_LUT25} that for all frequency values, the overclocked RCA achieves no larger error expectation than the RCA with truncated operand word-lengths, as predicted by the models for the RCA in Section~\ref{section_RCA_Comparison}. It can also be observed that the CSA cannot be implemented with the original word-length due to the limited area budget and this leads to large truncation errors. Although the CSA with 3 stages is best for some frequencies, the overclocked RCA is still the optimum design when high operating frequencies are applied.

We then perform similar experiments with a variety of area constraints. We implement CSA with all possible stage numbers within the given area specification. The optimum design methods with respect to different operating frequencies and area consumptions are demonstrated in Fig.~\ref{Fig_CSA_Tradeoff}, from which several observations can be made. Firstly, if the frequency requirement is not large whilst the available area is large enough to implement a CSA in full precision, it will be the optimum design. This is expected from our earlier analysis in Fig.~\ref{Fig_CSA2stage_Timing}. Secondly, the 2-stage CSA is better than the 3-stage CSA and the 4-stage CSA when frequency is low, as it consumes less area although all achieve the same error expectation. Thirdly, it is possible that only part of the CSA can be implemented under a tighter area budget, whereas the RCA still keeps full precision. In this case, area becomes the dominant factor and precision is lost for the CSA, meaning the RCA with overclocking is the optimum design method across almost the whole frequency domain in this situation. Finally, for very stringent area constraints, the word-length of RCA is also limited. This results in truncation errors initially for all design scenarios. However, the RCA with overclocking can still be employed as the optimum design, because it loses less precision than the CSA under the same area constraint.
%
\begin{figure}[t]
  \centering
  \vspace{-2.5ex}
  \includegraphics[width=4.5in]{./Figures/Tradeoff_Freq.eps}
  \vspace{-2.5ex}
  \caption{A demonstration of the optimum design methodology which achieves the minimum error at outputs with respect to a variety of frequency and area constraints.}
  \label{Fig_CSA_Tradeoff}
\end{figure}

\vspace{-0.5ex}
\subsection{Determination of the Optimum Adder Structure for Given Accuracy Requirements}
If the design goal is to operate the circuit as fast as possible with the minimum area whilst a certain error budget can be tolerated, the optimum design methodology can be decided as illustrated in Fig.~\ref{Fig_CSA_Tradeoff_Error}. In this situation, the error specifications are evaluated in terms of the mean relative error ($MRE$), as given by~(\ref{Eq:MRE}), where $E_{error}$ and $E_{out}$ refer to the mean value of error and the mean value of outputs, respectively.
%
\begin{eqnarray}\label{Eq:MRE}
  MRE=\left|\frac{E_{error}}{E_{out}}\right|\times 100\%
\end{eqnarray}

@@

@@
